%!TEX root = da2020.tex

This book is an introduction to the theory of distributed algorithms, with focus on distributed graph algorithms (network algorithms). The topics covered include:
\begin{itemize}
    \item \textbf{Models of computing:} precisely what is a distributed algorithm, and what do we mean when we say that a distributed algorithm solves a certain computational problem?
    \item \textbf{Algorithm design and analysis:} which computational problems can be solved with distributed algorithms, which problems can be solved \emph{efficiently}, and how to do it?
    \item \textbf{Computability and computational complexity:} which computational problems \emph{cannot} be solved at all with distributed algorithms, which problems cannot be solved \emph{efficiently}, why is this the case, and how to prove it?
\end{itemize}
No prior knowledge of distributed systems is needed. A basic knowledge of discrete mathematics and graph theory is assumed, as well as familiarity with the basic concepts from undergraduate-level courses on models on computation, computational complexity, and algorithms and data structures.


\usection{About the Course}

This textbook was written to support the lecture course \emph{CS-E4510 Distributed Algorithms} at Aalto University. The course is worth 5 ECTS credits. There are 12 weeks of lectures. Each week we will cover one chapter of this book, and our students are expected to solve the quiz and at least 3 of the exercises from the chapter.


\usection{Acknowledgments}

Many thanks to
Jaak\-ko Ala\-saari,
Alki\-da Balliu,
Sebastian Brandt,
Arthur Carels,
Jacques Charnay,
Faith Ellen,
Aelitta Ezugbaya,
Mika G\"o\"os,
Ja\-kob Greistorfer,
Nikos Heik\-kil\"a,
Joel Kaa\-si\-nen,
Samu Kal\-lio,
Mir\-co Kroon,
Tee\-mu Kuu\-sisto,
Dang Lam,
Tuo\-mo Lem\-pi\"a\-inen,
Christoph Len\-zen,
Darya Melnyk,
Abdul\-melik Mohammed,
Christopher Purcell,
Joel Ry\-bicki,
Joo\-na Sa\-ve\-la,
Stefan Schmid,
Roelant Steg\-mann,
Aleksandr Tereshchenko,
Przemys\l{}aw Uzna\'nski,
and
Jussi V\"ai\-s\"a\-nen
for feedback, discussions, comments, and for helping us with the arrangements of this course. This work was supported in part by the Academy of Finland, Grant 252018.


\usection{Website}

For updates and additional material, see
\begin{center}
    \url{https://jukkasuomela.fi/da2020/}
\end{center}


\usection{License}

\begin{samepage}
This work is licensed under the \emph{Creative Commons Attribution 4.0 International Public License}. To view a copy of this license, visit
\begin{center}
    \url{https://creativecommons.org/licenses/by/4.0/}
\end{center}
\end{samepage}
